---
title: "Enterotypes in the global Dataset"
author: "Marisa Keller"
output:
   html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(ggembl)
library(plotly)
library(ggpubr)
library(ggpattern)
library(ggExtra)
library(cowplot)
library(viridis)
library(PupillometryR) # for flat violine
library(markovchain)
library(igraph)
library(diagram)
library(ggsankey)

source(here("utils/Colors.R"))
theme_set(theme_publication())
outDir <- here("figures/")
```

# import data

```{r message=FALSE, warning=FALSE}
# taxonomic profiles
counts <- readRDS(here("data/motusMatrix_Genuslevel_gtdb_adults_filtered.rds"))
counts <- counts[row.names(counts)!="-1",]
counts <- na.omit(counts)
counts <- t(counts[rowSums(counts[, -1])>0,])

# metadata
taxonomy <- read_tsv(here("data/Taxonomy_gtdb.tsv"))
metadata <- readRDS(here("data/metadata.rds")) 

# cluster statistics
pam_stats <- readRDS(here("results/ClusterStatistics_PamClustering.rds"))
fkm_stats <- read_tsv(here("results/ClusterStatistics_FKMClustering.tsv"))

# rarafaction analysis
res_fkm <- read_tsv(file = here("results/Fuzzy_Rarefaction_Results.tsv"))
res_pam <- readRDS(here("results/Pam_rarefraction_results.rds"))

# dmm benchmarking 
dmmH <- read_tsv(here("results/DMM_benchmarking_Heterogenity.tsv"))
dmmS <- read_tsv(here("results/DMM_benchmarking_Size.tsv"))

# PCoA
pcoa.obj.adults <- readRDS(here("results/pcoa.rds")) 
```

# Correlation Bacteroides and Phocaicola

```{r}
correlation <-cor.test(counts[,"Bacteroides"], counts[,"Phocaeicola"], method = c("pearson"))
correlation
```
**Figure 1**

# fkm clustering 

```{r message=FALSE, warning=FALSE}

adults_filtered_gtdb <- res_fkm %>% filter(propStudies == 1) %>% 
  dplyr::rename(Silhouette = position) %>% 
  mutate(Silhouette = Silhouette +1) %>% 
  mutate(Silhouette = factor(Silhouette, levels = unique(Silhouette)),
         repetition = factor(repetition, levels = unique(repetition)))

df.plot <- adults_filtered_gtdb %>% 
  group_by(Silhouette) %>% 
  mutate(mean = mean(silhouette), 
         s.d. = stats::sd(silhouette), 
         upper = mean + s.d., 
         lower = mean - s.d., 
         n = length(unique(repetition))) %>% 
  filter(!is.na(repetition))


p_fkm_stat <- df.plot %>% 
  ggplot(aes(x=Silhouette, y = mean, group= repetition))+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.01, linetype = "dashed", color = "grey10")+
 geom_text(aes(x=5,y=0.5, label = "rep = 20", group = NA), size = 2,color = "grey10",  check_overlap = TRUE)+
  geom_line()+
  geom_vline(xintercept=2, color = "grey90", lty="dashed")+
   geom_point()+
  ylab("Silhouette")+xlab("# Clusters")
p_fkm_stat
```

## rarefy results

```{r}
rare.plot <- res_fkm %>% ungroup() %>% 
  mutate(propStudies=factor(propStudies)) %>%
  select(bestK_sil, propStudies, repetition) %>% distinct %>% 
  count(propStudies,bestK_sil) %>%
  ggplot(aes(x=propStudies,y=bestK_sil))+
  geom_tile(aes(fill=n))+
  scale_fill_gradient(low = "grey90",high="salmon")+
  geom_text(aes(label=n),color="grey10",alpha=0.5)+
  ylab("Best number of clusters (k)\nper iteration")+
  xlab("Proportion of studies used")+
  theme(legend.position = "bottom", legend.key.size = unit(0.5, "cm"))+
  labs(fill = "# of iterations")
rare.plot
```

## compare cohorts and taxonomy

```{r fig.height = 3, fig.width = 5}

res.s <- fkm_stats %>% 
  #dplyr::rename(Silhouette = position) %>% 
  mutate(Silhouette = word(Silhouette, 2,2, sep = fixed("=")) %>% as.numeric) %>% 
 # mutate(Silhouette = Silhouette +1) %>% 
  mutate(Silhouette = factor(Silhouette, levels = unique(Silhouette)),
         Rep = factor(Rep, levels = unique(Rep)))

df.plot <- res.s %>% 
  group_by(Silhouette, Dataset) %>% 
  mutate(mean = mean(value), 
         s.d. = stats::sd(value), 
         upper = mean + s.d., 
         lower = mean - s.d., 
         n = length(unique(Rep))) %>% 
  filter(!is.na(Rep))

# DATASETS
tmp <- df.plot %>% 
  dplyr::filter(Dataset %in% c("fkm_adults_filtered_gtdb", "fkm_adults_all_gtdb")) %>% 
  mutate(dataset = word(Dataset, 3,3, sep="_"),
         dataset = factor(dataset, levels = c("filtered", "all"))) %>% 
  ungroup() %>% 
  droplevels()
label <- tmp %>%  ungroup() %>%  select(Dataset, n) %>% distinct()
  
fkm_dataset <- 
  tmp %>% ggplot(aes(x=Silhouette, y = mean, group= Rep, color = Dataset, fill = Dataset))+
  geom_ribbon(aes(ymin = lower, ymax = upper, group = Dataset), alpha = 0.2, linetype = "dashed")+
  geom_line(aes(group = Dataset))+
  geom_vline(xintercept=2, color = "grey70", lty="dashed")+
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1),
        legend.position = "bottom")+
  xlab("Number of Cluster")+ylab("Silhouette Width [mean with sd]")+
  scale_color_manual(values = c("#16697a", "#00b4d8"))+
   scale_fill_manual(values = c("#16697a", "#00b4d8"))+
  geom_text(aes(x=7,y=0.5,label = paste("rep =", n), group = NA), size = 2,color = "grey10", data = label, check_overlap = TRUE)

fkm_dataset

# TAXONOMY 
tmp <- df.plot %>% 
  dplyr::filter(Dataset %in% c("fkm_adults_filtered_gtdb", "fkm_adults_filtered_ncbi")) %>% 
  mutate(taxonomy = word(Dataset, 4,4, sep = "_")) %>% 
  mutate(Dataset = word(Dataset, 2,4, sep="_"), 
         Dataset = factor(Dataset, levels = unique(Dataset))) %>% 
  ungroup() %>% 
  droplevels()
label <- tmp %>%  ungroup() %>%  select(taxonomy, n) %>% distinct()

fkm_taxonomy <- tmp %>% ggplot(aes(x=Silhouette, y = mean, group= Rep, color = taxonomy, fill = taxonomy))+
  geom_ribbon(aes(ymin = lower, ymax = upper, group = taxonomy), alpha = 0.2, linetype = "dashed")+
  geom_line(aes(group = taxonomy))+
  geom_vline(xintercept=2, color = "grey70", lty="dashed")+
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1),
        legend.position = "bottom")+
  xlab("Number of Cluster")+ylab("Silhouette Width [mean with sd]")+
 # facet_wrap(~taxonomy, ncol = 2)+
  scale_color_manual(values = c("#16697a", "#a3b18a"))+
  scale_fill_manual(values = c("#16697a", "#a3b18a"))+
  geom_text(aes(x=7,y=0.5,label = paste("rep =", n), group = NA), size = 2,color = "grey10", data = label, check_overlap = TRUE)

fkm_taxonomy
```


```{r fig.height = 2, fig.width =4}
# p.ch <- pam_stats[["ch_gtdb"]] %>% 
#   pivot_longer(cols = -nk, names_to = "dataset", values_to = "CH.index") %>% 
#   filter(dataset == "filtered") %>% 
#   ggplot(aes(x=nk, y = CH.index))+
#   geom_line()+
#   geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
#   #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
#  # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
#   geom_point()+
#   labs(title = "CH index")+
#   xlab("Number of Clusters")+ylab("")+
#   theme(legend.position = "blank")+
#   scale_x_continuous(n.breaks = 8)+
#   scale_color_manual(values = c("#7a9e9f","#4f6367",  "#b8d8d8"))
# 
# p.sil <-pam_stats[["sil_gtdb"]]  %>% 
#   pivot_longer(cols = -nk, names_to = "dataset", values_to = "Silhouette_Width") %>% 
#   filter(dataset == "filtered") %>% 
#   ggplot(aes(x=nk, y = Silhouette_Width))+
#   geom_line()+  geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
#   #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
#  # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
#   geom_point()+
#   theme(axis.title.y = element_blank(), legend.key.size = unit(0.2, "cm"))+
#   labs(title = "Sillhouette Width")+
#   xlab("Number of Clusters")+ylab("")+
#   scale_x_continuous(n.breaks = 8)+
#   scale_color_manual(values = c( "#7a9e9f","#4f6367", "#b8d8d8"))
# 
# ggarrange(p.ch, p.sil, nrow = 1)
```


# pam clustering 
## rarefy results

```{r}
# CH index 
rare.ch <- res_pam %>%
  dplyr::mutate(propStudies=factor(propStudies)) %>%
  count(propStudies,k_best_ch) %>%
  ggplot(aes(x=propStudies,y=k_best_ch))+
  geom_tile(aes(fill=n))+
  scale_fill_gradient(low = "grey90",high="salmon")+
  geom_text(aes(label=n),color="grey10",alpha=0.5)+
  theme_publication()+
  ylab("Best k per iteration")+
  xlab("Proportion of studies used")+
  theme(legend.position = "none", legend.key.size = unit(0.2, "cm"))+
labs(fill = "# of \niterations")
rare.ch

# Silhouette
rare.sil <- res_pam %>%
  mutate(propStudies=factor(propStudies)) %>%
  count(propStudies,k_best_sil) %>%
  ggplot(aes(x=propStudies,y=k_best_sil))+
  geom_tile(aes(fill=n))+
  scale_fill_gradient(low = "grey90",high="salmon")+
  geom_text(aes(label=n),color="grey10",alpha=0.5)+
  theme_publication()+
  theme(legend.key.size = unit(0.5, "cm"), legend.position = "bottom")+
  ylab("Best k per iteration")+
  xlab("Proportion of studies used")+
  labs(fill = "# of iterations")
rare.sil

```

## compare cohorts and taxonomy

```{r}
p_ch_gtdb <- pam_stats[["ch_gtdb"]] %>% 
  pivot_longer(cols = -nk, names_to = "dataset", values_to = "CH.index") %>% 
  filter(dataset != "healthy") %>% 
  ggplot(aes(x=nk, y = CH.index, color = dataset))+
  geom_line()+
  geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
  #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
 # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
  geom_point()+
  theme_publication()+
  #labs(title = "CH index")+
  xlab("Number of Clusters")+ylab("CH-Index")+
  theme(legend.position = "blank")+
  scale_x_continuous(breaks = 1:20)+
  scale_color_manual(values = c("#00b4d8","#4f6367",  "#b8d8d8"))
p_ch_gtdb

p_ch_gtdb_filtered <- pam_stats[["ch_gtdb"]] %>% 
  pivot_longer(cols = -nk, names_to = "dataset", values_to = "CH.index") %>% 
  filter(dataset != "healthy") %>% 
  filter(dataset == "filtered") %>% 
  ggplot(aes(x=nk, y = CH.index, color = dataset))+
  geom_line()+
  geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
  #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
 # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
  geom_point()+
  theme_publication()+
  #labs(title = "CH index")+
  xlab("Number of Clusters")+ylab("CH-Index")+
  theme(legend.key.size = unit(0.2, "cm"))+
  scale_x_continuous(breaks = 1:20)+
  scale_color_manual(values = c("#4f6367",  "#b8d8d8"))
p_ch_gtdb_filtered

p_sil_gtdb <- pam_stats[["sil_gtdb"]]  %>% 
  pivot_longer(cols = -nk, names_to = "dataset", values_to = "Silhouette_Width") %>% 
  filter(dataset != "healthy") %>% 
  ggplot(aes(x=nk, y = Silhouette_Width, color = dataset))+
  geom_line()+  geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
  #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
 # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
  geom_point()+
  theme_publication()+
  theme( legend.key.size = unit(0.2, "cm"))+
  #labs(title = "Sillhouette Width")+
  xlab("Number of Clusters")+ylab("Silhouette Width")+
  scale_x_continuous(breaks = 1:20)+
  scale_color_manual(values = c( "#00b4d8","#4f6367", "#b8d8d8"))
p_sil_gtdb

p_sil_gtdb_filtered <- pam_stats[["sil_gtdb"]]  %>% 
  pivot_longer(cols = -nk, names_to = "dataset", values_to = "Silhouette_Width") %>% 
  filter(dataset != "healthy") %>% 
  filter(dataset == "filtered") %>% 
  ggplot(aes(x=nk, y = Silhouette_Width, color = dataset))+
  geom_line()+  geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
  #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
 # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
  geom_point()+
  theme_publication()+
  theme(legend.key.size = unit(0.2, "cm"))+
  #labs(title = "Sillhouette Width")+
  xlab("Number of Clusters")+ ylab("Silhouette Width")+
  scale_x_continuous(breaks = 1:20)+
  scale_color_manual(values = c( "#4f6367", "#b8d8d8"))
p_sil_gtdb_filtered

p_ch_ncbi <- pam_stats[["ch_ncbi"]]%>% 
  filter(nk != 1) %>% 
  pivot_longer(cols = -nk, names_to = "dataset", values_to = "CH.index") %>% 
  ggplot(aes(x=nk, y = CH.index, color = dataset))+
  geom_line()+ geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
  #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
 # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
  geom_point()+
  theme_publication()+
  theme(legend.position = "blank")+
  scale_x_continuous(breaks = 1:20)+
  xlab("Number of Clusters")+ylab("CH-Index")+
  scale_color_manual(values = c("#4f6367","#a3b18a",  "#b8d8d8"))
p_ch_ncbi

p_sil_ncbi <- pam_stats[["sil_ncbi"]] %>% 
  filter(nk != 1) %>% 
  pivot_longer(cols = -nk, names_to = "dataset", values_to = "Silhouette.width") %>% 
  ggplot(aes(x=nk, y = Silhouette.width, color = dataset))+
  geom_line()+ geom_vline(xintercept = 3, color = "grey70", lty = "dashed")+
  #geom_vline(xintercept = 3, color = "grey60", lty="dashed")+
 # geom_vline(xintercept = 2, color = "grey80", lty="dashed")+
  geom_point()+
  theme_publication()+ 
  labs(color="taxonomy")+
  theme( legend.key.size = unit(0.2, "cm"))+
  scale_x_continuous(breaks = 1:20)+
  xlab("Number of Clusters")+ylab("Silhouette Width")+
  scale_color_manual(values = c("#4f6367","#a3b18a", "#ee9b00"))
p_sil_ncbi
```

Supplementary Fig. 1: CH-index and Silhouette width for different Cohorts (A-B) and for clustering based on different genus-level taxonomies (C-D)

# PCoA
## fkm 
k=3 fkm Cluster Assignments
```{r}
summary(pcoa.obj.adults)
```

```{r fig.height = 5.5, fig.width = 5.5}

p <- ggplot(metadata, aes(x=A1, y=A2, color = fkm_k3, shape = fkm_k3))+
  geom_hline(yintercept = 0, color = "grey50")+
  geom_vline(xintercept = 0, color = "grey50")+
  geom_point(size = 0.5, 
             #shape = 21
             alpha = 0.5)+
  annotate(geom="text", x = 0.17, y = 0.23, label ="Firmicutes ET", color = "#197EA5", fontface = "bold", size = 3)+
  annotate(geom="text", x = 0.16, y = -0.4, label ="Bact_Phoc ET", color = "#2B9348", fontface = "bold", size = 3)+
 annotate(geom="text", x = - 0.3, y = -0.4, label ="Prevotella ET", color = "#DF6929", fontface = "bold", size = 3)+ # size = 3 for publication
  #labs(subtitle = "PCoA | FKM Clustering | k=3")+
  xlab("Axis 1 [15.47 %]")+ # find out about the percentage by "summary(pcoa.obj)"
  ylab("Axis 2 [9.01 %]")+
  scale_color_manual(values = enterotype_colors, breaks = names(enterotype_colors))+
  scale_shape_manual(values = enterotype_shapes, breaks = names(enterotype_shapes))+
  theme(legend.position = "none", axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank(),
        #axis.title = element_text(size = 10),
        plot.subtitle = element_text(vjust = -0.5, size = 12)
       )
  
p3 <- ggMarginal(p, groupColour = TRUE, groupFill = TRUE, alpha =.3)

p3
```

```{r fig.width = 8, fig.height=1}
all_samples_ET <- metadata %>% 
  count(fkm_k3) %>% 
  unite("text", fkm_k3, n, sep = " \n n =", remove = FALSE) %>% 
  ggplot(aes(x=1, y = n, fill = fkm_k3, label = text))+
  geom_bar(position="stack", stat="identity", alpha = 0.8)+
  coord_flip()+
  theme_void()+
  theme(legend.position = "none")+
  scale_fill_manual(values = enterotype_colors, breaks = names(enterotype_colors))+
  geom_text(size = 3, position = position_stack(vjust = 0.5))
  
all_samples_ET
```

## pam 
k=3 Cluster Assignments

```{r}
p <- ggplot(metadata, aes(x=A1, y=A2, color = pam_k3))+
  geom_hline(yintercept = 0, color = "grey50")+
  geom_vline(xintercept = 0, color = "grey50")+
  geom_point(size = 0.5, shape = 21,  alpha = 0.5)+
 # labs(title = "PCoA | PAM Clustering | k=3")+
  annotate(geom="text", x = 0.15, y = 0.23, label ="Firmicutes ET", color = "#197EA5", size = 2)+
  annotate(geom="text", x = 0.16, y = -0.4, label ="Bact_Phoc ET", color = "#2B9348", size = 2)+
  annotate(geom="text", x = - 0.3, y = -0.4, label ="Prevotella ET", color = "#DF6929", size = 2)+
  xlab("Axis 1 [15.47 %]")+ # find out about the percentage by "summary(pcoa.obj)"
  ylab("Axis 2 [9.01 %]")+
  theme_publication()+
  scale_color_manual(values = enterotype_colors, breaks = names(enterotype_colors))+
  theme(legend.position = "none", axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank())
  
pam.k3 <- ggMarginal(p, groupColour = TRUE, groupFill = TRUE, alpha =.3)

pam.k3
```

k=2 Cluster Assignments

```{r}
p <- ggplot(metadata, aes(x=A1, y=A2, color = pam_k2))+
  geom_hline(yintercept = 0, color = "grey50")+
  geom_vline(xintercept = 0, color = "grey50")+
  geom_point(size = 0.5, shape = 21,  alpha = 0.5)+
  #labs(title = "PCoA | PAM Clustering | k=2")+
  annotate(geom="text", x = 0.18, y = 0.25, label ="Bact_Phoc /\nFirmicutes ET", color = "#2B9348", size =2 )+
  annotate(geom="text", x = - 0.3, y = -0.4, label ="Prevotella ET", color = "#DF6929", size = 2)+
  theme_publication()+
  xlab("Axis 1 [15.47 %]")+ # find out about the percentage by "summary(pcoa.obj)"
  ylab("Axis 2 [9.01 %]")+
  scale_color_manual(values = enterotype_colors, breaks = names(enterotype_colors))+
  theme(legend.position = "none", 
        axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank())
  
pam.k2 <- ggMarginal(p, groupColour = TRUE, groupFill = TRUE, alpha =.3)

pam.k2
```

# DDA of genera between ET

## calculate DDA 

first: define prevalence cutoff
we plot the number of samples per taxa to define a suitable prevalence cut-off
```{r}
df <-tibble(prevalence = colSums(counts != 0), 
       taxa = colnames(counts)) %>% 
  arrange(desc(prevalence)) %>% 
  mutate(taxa = factor(taxa, levels = taxa))

df %>% 
  ggplot(aes(x=taxa,y = prevalence))+
  geom_col()+
  theme(axis.text.x = element_blank())+
  geom_hline(yintercept = 100, color = "orange", lty = "dashed")

```
cutoff for prevalence to 100.
The taxa needs to be in at least 100 samples of the dataset; this is equivalent to 0.5 % of the samples 
```{r}
data_dda <- counts[,colnames(counts) %in% (df %>% dplyr::filter(prevalence >= 101) %>% pull(taxa))]
```
get only one sample per patient
```{r}
first_visit <- aggregate(metadata$internal_id, by =list(metadata$subject_id), FUN=dplyr::first) %>% 
  magrittr::set_colnames(c("subject_id", "internal_id"))

data_dda <- data_dda[rownames(data_dda) %in% first_visit$internal_id,]
md_dda <- metadata %>% filter(internal_id %in% first_visit$internal_id)
```

Calculate the Wilcoxon Test
```{r eval=FALSE, include=FALSE}
source(here("utils/WilcoxonTest.R"))
log.n0 = 1e-05 #species
log.n0.func = 1e-08 #functions

df.BP <- calWilcox(data_dda %>% t(), 
          md_dda %>% dplyr::mutate(ID = internal_id) %>% tibble::column_to_rownames("internal_id") %>% 
            dplyr::rename(status = fkm_k3), 
          "DAA_BPvsOthers_k3", 
          resultfolder = here::here("results/"),
          case="Bact_Phoc", ctr=c("Prevotella", "Firmicutes_1"))

df.F <- calWilcox(data_dda %>% t(), 
          md_dda %>% dplyr::mutate(ID = internal_id) %>% tibble::column_to_rownames("internal_id")  %>% 
          dplyr::rename(status = fkm_k3), 
          "DAA_FvsOthers_k3", 
          resultfolder = here::here("results/"),
          case="Firmicutes_1", ctr=c("Prevotella", "Bact_Phoc")) 

df.P <- calWilcox(data_dda %>% t(), 
          md_dda %>% dplyr::mutate(ID = internal_id) %>% tibble::column_to_rownames("internal_id") %>% 
          dplyr::rename(status = fkm_k3), 
          "DAA_PvsOthers_k35", 
          resultfolder = here::here("results/"),
          case="Prevotella", ctr=c("Firmicutes_1", "Bact_Phoc")) 

```

summarize the results
```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
df.dda <- bind_rows(df.BP %>% dplyr::mutate(enterotype = "Bact_Phoc") , 
                    df.F %>% dplyr::mutate(enterotype = "Firmicutes"), 
                    df.P %>% dplyr::mutate(enterotype = "Prevotella"))



df.plot <- df.dda %>% 
  arrange(desc(rel)) %>% 
  mutate(species = factor(species, levels = unique(species))) %>%
  filter(rel >= 0.05)

# save the results
#saveRDS(df.dda, here("results/DDA_betweenET.rds"))


```

## plot DDA
```{r}
dda <- readRDS(here("results/DDA_betweenET.rds"))

taxonomy <- taxonomy %>% select(-c(onlyID, species)) %>% distinct()

dda <- dda %>% left_join(taxonomy, by = c("species" = "genus")) %>% 
  dplyr::mutate(phylum = case_when(species == "Acetatifactor.Coprococcus" ~ "Lachnospiraceae",
                            TRUE ~ phylum))

# adapt some species names 
dda <- dda %>% 
  mutate(species = case_when(species == "CAG.83" ~ "unclassified Oscillospiraceae", 
                             species == "Lachnospira.CAG.196" ~ "unclassified Lachnospira",
                                TRUE ~ species),
         phylum = case_when(species == "unclassified Oscillospiraceae" ~ "Firmicutes", 
                             species == "unclassified Lachnospira" ~ "Firmicutes",
                                TRUE ~ phylum))
enterotypes <- unique(dda$enterotype)
```

```{r fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
dda_plots <- list()
for(e in 1:length(enterotypes)) {
  et <- enterotypes[[e]]
  print(et)
 p <-  dda %>% 
  filter(enterotype == et) %>%
  filter(adj < 0.01) %>%   filter(rel >= 0.1) %>% 
  arrange(fc) %>% 
  mutate(species = factor(species, levels =unique(.$species))) %>% 
  ggplot(aes(x=fc, y = species, fill  = phylum)) +
    geom_col(position = "dodge")+
    labs(subtitle = paste(et, "ET"))+
   xlab("fold change")+
   scale_fill_manual(values = phyla_colors)+
   theme(legend.position = "none", 
         axis.title.y = element_blank())
 print(p)
 dda_plots[[e]] <- p
}

legend <- get_legend(ggplot(taxonomy, aes(fill= phylum))+scale_fill_manual(values=phyla_colors))
dda
```


# longitudinal Analysis

```{r}
# Filter metadata for subjects with multiple timepoints
md_time <- metadata %>% filter(!is.na(subject_id)) %>% 
  # for each subject, count how many samples we have and filter for subjects with >= 2 samples 
  group_by(subject_id) %>% 
  mutate(n_samples_per_subject = n()) %>% 
  ungroup() %>% 
  dplyr::filter(n_samples_per_subject >=2) %>% 
  dplyr::filter(!is.na(timepoint)) %>% 
  # for each sample calculate the time between the sample and the first sample and remove everything above 365 days 
  group_by(subject_id) %>% 
  dplyr::mutate(first.visit = min(collection_date),
                time.from.first.visit = collection_date - first.visit) %>% 
  ungroup() %>% 
  dplyr::filter(time.from.first.visit <= 365)

tmp <- md_time %>% 
  group_by(subject_id) 

md_time %>% ungroup() %>% 
  select(subject_id, n_samples_per_subject, studies) %>% distinct() %>% 
  filter(n_samples_per_subject >= 6) %>% count(studies)

samplesPerStudy.plot <- md_time %>% select(subject_id, n_samples_per_subject) %>% distinct() %>% count(n_samples_per_subject) %>% 
  ggplot(aes(x=n_samples_per_subject, y = n))+
    geom_line()+
  geom_point(shape = 21, size = 4, fill = "white")+
  geom_point(aes(x=89, y = 1), shape = 4, size = 4,color = "red")+
  scale_x_continuous(n.breaks = 10)+  scale_y_continuous(n.breaks = 10)+
  ylab("Number of Subjects")+xlab("Number of Samples per Subject")
samplesPerStudy.plot
```

## Markov Chain Model

input to markov chain model. a list of vectors with the enterotype number ordered by the timepoints for each patient.

```{r}
et_vectorlist <- md_time %>%
  # remove the outlier with 90 samples to not get biased by this
  filter(subject_id != "Fukuyama_2017_bowelcleansing_AAB") %>% 
 select(subject_id, timepoint, fkm_k3) %>% 
  dplyr::mutate(pam_k3 = case_when(fkm_k3 == "Firmicutes" ~ "Firmicutes", 
                   TRUE ~ fkm_k3)) %>% 
  group_by(subject_id, .add = TRUE) %>% 
  arrange(timepoint, .by_group = TRUE) %>% 
  select(-timepoint) %>% 
  group_by(subject_id, .add=TRUE) %>% 
     group_split(subject_id) %>% 
     map(~ .x %>%
             pull(fkm_k3))

single_value_vectors <- sapply(et_vectorlist, function(x) length(unique(x)) == 1)

```

calulating transition matrix

```{r message=FALSE, warning=FALSE}
mcFit <- markovchainFit(et_vectorlist, method = "mle")

mcDf <- as(mcFit$estimate, "data.frame") 
```

Plottin the Markov Chain Model

```{r eval=FALSE, fig.height=4, fig.width=4, include=FALSE}
tM <- mcFit$estimate@transitionMatrix %>% round(., digits = 2) %>% t()
tM %>% as.matrix() %>% colSums
tM <- tM[c("Bact_Phoc", "Prevotella", "Firmicutes"), c("Bact_Phoc", "Prevotella", "Firmicutes")] 
colnames(tM) <- c("B/P ET", "P ET", "F ET")
rownames(tM) <- c("B/P ET", "P ET", "F ET")

png(file=here("figures/MarkovChain.png"), width = 13, height =12)
diagram::plotmat(tM, curve=.05,
                 relsize=0.5,
                 arr.type = "simple",
                 box.type = "rect",
                 box.prop = 0.7,
                 box.lwd = 1,
                 box.size = 0.15,
                 box.cex = 1.8,
                 box.col = c("#2B9348", "#DF6929", "#197EA5"),
                 shadow.size= 0,
                 self.shiftx = c(0.23,-0.22,0),
                 self.shifty = c(0, 0, 0.17),
                 arr.lwd =tM*10,
                 dtext = 0.21,
                 arr.width = tM*10,
                 self.lwd = diag(tM)*10, 
                 arr.pos=0.5,
                 cex = 1.8)
dev.off()
```

## analysing the changes

```{r}
md_time_changes <- md_time %>%
  # remove the outlier with 90 samples to not get biased by this
  filter(subject_id != "Fukuyama_2017_bowelcleansing_AAB") %>% 
  select(subject_id, timepoint, fkm_k3, internal_id, ET_dysbiosis, max_prob) %>% 
  dplyr::mutate(fkm_k3 = case_when(fkm_k3 == "Firmicutes" ~ "Firmicutes", 
                   TRUE ~ fkm_k3)) %>% 
  group_by(subject_id, .add = TRUE) %>% 
  arrange(timepoint, .by_group = TRUE) %>% 
  group_by(subject_id, .add =TRUE) %>% 
  mutate(next_fkm_k3 = lead(fkm_k3)) %>% # document the next enterotype
  mutate(change = fkm_k3 != lead(fkm_k3)) %>%  # create new variable "change"
  filter(!is.na(next_fkm_k3)) %>% 
  unite(ET_change, fkm_k3, next_fkm_k3, remove = FALSE, sep = "_to_") %>% 
  mutate(ET_change = factor(ET_change, levels = c("Bact_Phoc_to_Firmicutes", "Bact_Phoc_to_Prevotella", "Bact_Phoc_to_Bact_Phoc",
                                                  "Prevotella_to_Bact_Phoc", "Prevotella_to_Firmicutes","Prevotella_to_Prevotella",
                                                   "Firmicutes_to_Bact_Phoc", "Firmicutes_to_Prevotella", "Firmicutes_to_Firmicutes"))) %>% 
  mutate(change = case_when(change == TRUE ~ "change", 
                             change == FALSE ~ "no change")) %>% 
  ungroup() %>%  group_by(ET_change) %>% 
  mutate(number_in_each_change = n())
```

## longitudinal Examples in PCoA

```{r}
time_subjects <- md_time %>% filter(n_samples_per_subject >= 10) %>%
  select(n_samples_per_subject, subject_id) %>% distinct() %>% pull(subject_id)
```

```{r, fig.width = 4, fig.height = 4}
plot.list <- list()
for(s in 1:length(time_subjects)){
  subject <- time_subjects[[s]]
p <- ggplot(metadata %>% filter(subject_id != subject), aes(x=A1, y=A2, color = fkm_k3))+
  geom_hline(yintercept = 0, color = "grey50")+
  geom_vline(xintercept = 0, color = "grey50")+
  geom_point(size = 0.5, shape = 21, alpha = 0.2)+
  geom_path(data = metadata %>% filter(subject_id == subject), aes(x=A1, y=A2), color = "grey30")+
  geom_point(data = metadata %>% filter(subject_id == subject) %>% arrange(timepoint),aes(x=A1, y=A2, fill = fkm_k3), shape = 21,color = "grey30")+
  
  annotate(geom="text", x = 0.15, y = 0.23, label ="Firmicutes ET", color = "#197EA5", fontface = "bold")+
  annotate(geom="text", x = 0.16, y = -0.4, label ="Bact_Phoc ET", color = "#2B9348", fontface = "bold")+
  annotate(geom="text", x = - 0.3, y = -0.4, label ="Prevotella ET", color = "#DF6929", fontface = "bold")+
  labs(subtitle = paste(subject, "\nPCoA k=3 | FKM Clustering"))+
  xlab("Axis 1 [15.47 %]")+ # find out about the percentage by "summary(pcoa.obj)"
  ylab("Axis 2 [9.01 %]")+
  scale_color_manual(values = enterotype_colors, breaks = names(enterotype_colors))+
  scale_fill_manual(values = enterotype_colors, breaks = names(enterotype_colors))+
  theme(legend.position = "none", axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank(),
        axis.title = element_text(size = 10))

p
plot.list[[s]] <- p
}

plot.list
```

# Enterotypes in single Studies

```{r}
ET_count <- metadata %>% 
  group_by(studies) %>% 
  mutate(SampleNumber = n()) %>% 
  filter(SampleNumber >= 9) %>% 
  summarise(et_count= n_distinct(fkm_k3))
```

```{r, fig.width=15, fig.height = 20}
df <- metadata %>% 
  filter(!is.na(fkm_k3)) %>% 
  # calculate number of samples per study
  select(studies, fkm_k3) %>% 
  group_by(studies) %>% 
  mutate(SampleNumber = n()) %>% 
  # calculate proportion of samples in each enterotype
  filter(SampleNumber >= 9) %>% 
  group_by(studies, SampleNumber) %>% 
  count(fkm_k3) %>% 
  mutate(n_prob = n/SampleNumber) 
  
df <- df %>% 
  left_join(ET_count, by = "studies")

p1 <- df %>% 
  mutate(fkm_k3 = factor(fkm_k3, levels = c("Firmicutes", "Bact_Phoc", "Prevotella"))) %>%  # arrange for order by Firmicutes
  group_by(fkm_k3) %>% 
  arrange(desc(n_prob), .by_group = TRUE) %>% 
  mutate(studies = factor(studies, levels = unique(.$studies))) %>% 
  mutate(fkm_k3 = factor(fkm_k3, levels = c("Prevotella", "Firmicutes", "Bact_Phoc"))) %>%  # arrange for order in Plot
  ggplot(aes(x=n_prob, y = studies, fill = fkm_k3))+
  geom_col(position = "stack", color = "white")+
  scale_fill_manual(name= "enterotype", values = enterotype_colors)+
  theme(legend.position = "top")+
  xlab("Proportion of samples in each enterotype")
p2 <- df %>% 
  mutate(ET_count = as.factor(et_count)) %>% 
  ggplot(aes(x=studies, y = "n", fill = ET_count))+
  geom_tile(color = "white")+
  scale_fill_manual(values = c("#E76F51", "#2A9D8F", "#F4A261", "#52796F"))+
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank(), 
        legend.position = "right", 
        legend.title = )+
  coord_flip()

p3.studies <- df %>% 
  mutate(ET_count = as.factor(et_count)) %>% 
 mutate(fkm_k3 = factor(fkm_k3, levels = c("Firmicutes", "Bact_Phoc", "Prevotella"))) %>%  # arrange for order by Firmicutes
  group_by(fkm_k3) %>% 
  arrange(desc(n_prob), .by_group = TRUE) %>% 
  mutate(studies = factor(studies, levels = unique(.$studies))) %>% 
  ungroup() %>% 
  select(studies, SampleNumber) %>% distinct() %>% 
  ggplot(aes(x=studies, y = SampleNumber))+
  geom_col()+
  geom_text(aes(label = SampleNumber),position = position_stack(vjust = 1), hjust = -0.1, size = 3)+
   theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank(), 
        legend.position = "right", 
        legend.title = )+
  ylab("Sample number in each study")+
  coord_flip()
 


study.p <- ggarrange(p1,p3.studies,  ncol = 2, widths = c(9,9), align = "h", common.legend = TRUE)
study.p
```
proportion of enterotype classifications in each study 
```{r}
df.total <- metadata %>% 
  mutate(n_samples_total = n()) %>% 
  group_by(fkm_k3, n_samples_total) %>% 
  count(fkm_k3) %>% 
  mutate(sample_proportion_total = n/n_samples_total)
  
study.bp.df <- df %>% 
  filter(n>=9) %>% 
  group_by(studies) %>% 
  mutate(max_prob = max(n_prob))

study.bp<-study.bp.df %>% 
ggplot(aes(x=fkm_k3, y = n_prob, fill = fkm_k3))+
  geom_boxplot(outlier.color = "grey70")+
  geom_point(data = df.total, aes(x=fkm_k3, y = sample_proportion_total), shape = 23, size = 2, fill = "grey80")+
# geom_jitter(shape =21, aes(fill = as.factor(et_count)), width = .2, size = 1.2, alpha = 0.6)+
  stat_compare_means(comparisons = list(c("Bact_Phoc", "Firmicutes"), c("Bact_Phoc", "Prevotella"), c("Firmicutes", "Prevotella")), 
                     method = "wilcox.test", 
                     label = "p.signif",
                     vjust = 0.7,
                     step.increase = 0.07)+
  #scale_fill_manual(values = c("red", "blue", "grey80"))+
  scale_fill_manual(values = enterotype_colors)+
  xlab("")+ylab("proportion of samples per study")+
  labs(fill = "# of \nassigned \nEnterotypes")+
  theme(legend.position = "none", legend.key.size = unit(0.2, "cm"), axis.text.x = element_text(angle=45, hjust=1))
study.bp
```

# Functional MDS
```{r}
# read data 
comp_enterotype <- readRDS(here("results/keggko_enterotype_qc_rel_sub_sqrt_PCA_data.rds"))
# Plot MDS dimensions
## dim 1 & 2
mds12 <- ggplot(comp_enterotype, aes(x = Dim2, y = Dim1)) +
  geom_point(aes(color = fkm_k3), alpha = .3, size = 1) +
  theme_minimal() +
  
  scale_color_manual(values = enterotype_colors )+
  theme_publication()+
  ylab("Dim 1 [45%]")+xlab("Dim 2 [28%]")+
  theme(
    legend.position = "none",
  ) 
## dim 1 & 3
mds13 <- ggplot(comp_enterotype, aes(x = Dim3, y = Dim1)) +
  geom_point(aes(color = fkm_k3), alpha = .3, size = 1) +
  theme_minimal() +
  ylab("Dim 1 [45%]")+xlab("Dim 3 [12%]")+
  scale_color_manual(values = enterotype_colors )+
  theme_publication()+
    theme(
    legend.position = "none"
  ) 

```

# Alluvial Diagramm

```{r}
library("ggalluvial")
pallu <- metadata %>%
  filter(disease == "None reported") %>%
  dplyr::count(fkm_k4, pam_k4, fkm_k3, pam_k3, fkm_k2, pam_k2,
               #dmm_k3
               ) %>% as.data.frame() %>%
  mutate_if(is.double, as.character)
```

```{r}
pam <- pallu %>% 
  #mutate(pam_k3 = factor(pam_k3, levels = c("Bact_Phoc", "Firmicutes", "Prevotella"))) %>% 
  ggplot(aes(axis1 = pam_k2, axis2 = pam_k3, axis3 = pam_k4, y = n))+
 geom_stratum(width = 1/3,alpha =0.6 )+
  stat_alluvium(aes(fill=pam_k3),
                aes.bind = "flows",
                lode.guidance = "zigzag")+
  geom_alluvium(aes(fill=pam_k3))+
  geom_text(stat="stratum", aes(label = after_stat(stratum)), angle = 90, size = 3)+
  #stat_alluvium(lode.guidance = "zigzag")+
  scale_x_discrete(limits=c("k=2", "k=3", "k=4"))+
  scale_fill_manual(values = enterotype_colors)+
   theme_minimal()+
#  labs(title = "Enterotype Classification \nfor each sample \nPAM Clustering")+
  theme(legend.position = "none",panel.grid = element_line(color = "grey80", 
        size = 0.5), panel.grid.minor = element_blank(), 
        axis.title.y = element_text(size = 8))+
  ylab("Number of Samples")
pam

alluvial_pamfuzzy <- pallu %>% ggplot(aes(axis1 = fkm_k3, axis2 = pam_k3 , y = n))+
  geom_stratum(width = 1/3,alpha =0.6 )+
  stat_alluvium(aes(fill=fkm_k3),
                aes.bind = "flows",
                lode.guidance = "zigzag")+
  geom_alluvium(aes(fill=fkm_k3))+
  geom_text(stat="stratum", aes(label = after_stat(stratum)), angle = 90, size = 2)+
  #stat_alluvium(lode.guidance = "zigzag")+
  scale_x_discrete(limits=c( "FKM", "PAM" ))+
  #theme_minimal()+
  theme_publication()+
  ylab("Number of Samples")+
  scale_fill_manual(values = enterotype_colors)+
  #labs(title = "Enterotype Classification \nfor each sample")+
  theme(legend.position = "none",panel.grid = element_line(color = "grey80", 
        size = 0.5), panel.grid.minor = element_blank(), 
        axis.title.y = element_text(size = 8))
  
   alluvial_pamfuzzy
```


# Comparing with Frioux 2023 and Tap 2023
```{r}
comp.data <- readRDS(here("data/Entertoypes_Frioux_Tap_mapping.rds"))
```

```{r}
alpha.vector <- vctrs::vec_rep_each(comp.data %>% 
  filter(!is.na(branch)) %>% filter(!is.na(ES)) %>% pull(ET_dysbiosis),3)
sankey.all <- comp.data %>% 
  filter(!is.na(branch)) %>% filter(!is.na(ES)) %>% 
  dplyr::rename(Enterotypes = fkm_k3, `Tap et al. 2023` = branch, `Frioux et al. 2023` = ES) %>% 
  make_long(`Tap et al. 2023`,Enterotypes, `Frioux et al. 2023`)
sankey.plot <- sankey.all %>%

ggplot(
       aes(x = x, 
               next_x = next_x, 
               node = node, 
              next_node = next_node, 
               fill = factor(node),
               label = node)) +
  geom_sankey(flow.alpha = 0.5,
              node.color = 1, na.rm = TRUE) +
  geom_sankey_label(size = 3, color = 1, fill = "white", alpha = 0, hjust = 1.25) +
  theme_sankey(base_size = 12)+
  scale_alpha_identity()+
  scale_fill_manual(values = c("#ffd166", "#2B9348","#2B9348","#ffd166",  "#197EA5", "#197EA5","#DF6929",
                               "#2B9348","#2B9348","#2B9348","#197EA5",
                                "#DF6929", "#DF6929", "#197EA5","#197EA5"))+
  theme(legend.position = "none")+
  xlab("")

sankey.plot
```



```{r}
alpha.vector <- vctrs::vec_rep_each(comp.data %>% pull(ET_dysbiosis),3)
sankey.all <- comp.data %>% 
  dplyr::rename(Enterotypes = fkm_k3, `Tap et al. 2023` = branch, `Frioux et al. 2023` = ES) %>% 
  make_long(`Tap et al. 2023`,Enterotypes, `Frioux et al. 2023`)
sankey.all.plot <- sankey.all %>%

ggplot(
       aes(x = x, 
               next_x = next_x, 
               node = node, 
              next_node = next_node, 
               fill = factor(node),
               label = node)) +
  geom_sankey(flow.alpha = 0.5,
              node.color = 1, na.rm = TRUE) +
  geom_sankey_label(size = 3, color = 1, fill = "white", alpha = 0, hjust = 1.25) +
  theme_sankey(base_size = 12)+
  scale_alpha_identity()+
  scale_fill_manual(values = c("#ffd166", "#2B9348","#2B9348","#ffd166",  "#197EA5", "#197EA5","#DF6929",
                               "#2B9348","#2B9348","#2B9348","#197EA5",
                                "#DF6929", "#DF6929", "#197EA5","#197EA5"))+
  theme(legend.position = "none")+
  xlab("")

sankey.all.plot
```

```{r}
Dysbiosis.change <- comp.data %>% 
  filter(!is.na(branch)) %>% filter(!is.na(ES)) %>% 
  mutate(Change = case_when(type.change.x == "change" & type.change.y == "change" ~ "3 types",
                            type.change.x == "change" ~ "2 types", 
                            type.change.y == "change" ~ "2 types",
                            type.change.x == "no change" & type.change.y == "no change" ~ "1 type", 
                            TRUE ~as.character(NA))) %>% 
  mutate(Change = factor(Change, levels = c("1 type", "2 types", "3 types"))) %>% 
  ggplot(aes(y=max_prob, x = Change, fill = Change))+
  geom_boxplot(color= "grey20", varwidth = TRUE, alpha = 0.5,position= position_dodge(width = 0.2))+
  scale_fill_manual(values = c("#ffce1f","#D17EC3", "#aa3e98"))+
  ggembl::theme_publication()+
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust =1))+ 
  stat_compare_means(comparisons = list(c("1 type", "3 types"), 
                                        c("1 type", "2 types"), 
                                        c("3 types", "2 types")), aes(label = after_stat(p.signif)), vjust = 0.4)+
  ylab("classification strength")+xlab("sample classified into")

Dysbiosis.change

```