---
title: "Enterotype Dysbiosis Score"
author: "Marisa Keller"
output:
   html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(ggembl)
library(ggridges)
library(PupillometryR)
library(ggpubr)
library(viridis)

theme_set(theme_publication())
source(here("utils/Colors.R"))
```

import data
```{r message=FALSE, warning=FALSE}
metadata <- readRDS(here("data/metadata.rds")) 
counts <- readRDS(here("data/motusMatrix_Genuslevel_gtdb_adults_filtered.rds"))
# add dysbiosis score calculated by Lloyd-Price et al. 2019
dScore <- read_tsv(here("data/LP_dysbiosis_index_data.tsv"))
```

define disese and healthy group
```{r}
metadata <- metadata %>% 
  left_join(dScore, by = "internal_id") %>% 
    mutate(is.diseased = case_when(disease == "None reported" ~ "healthy subjects", 
                                 disease != "None reported" ~ "patients"),
           is.diseased = case_when(is.na(disease) ~ "unknown disease status", TRUE ~ is.diseased),
           fkm_k3 = factor(fkm_k3, levels = c("Prevotella", "Firmicutes", "Bact_Phoc"))) #%>% 
  # dplyr::mutate(is.diseased = factor(is.diseased, levels = c("healthy subjects","unknown disease status", "patients"))) %>% 
  # dplyr::mutate(fkm_k3_dys = case_when(is.dysbiosis == "dysbiosis" ~ "dysbiosis", TRUE ~ fkm_k3))
```


# Enterotype Dysbiosis Score

## PCoA with classification strength 
```{r}
pprob <- ggplot(metadata, aes(x=A1, y=A2, color = max_prob))+
  geom_hline(yintercept = 0, color = "grey50")+
  geom_vline(xintercept = 0, color = "grey50")+
  geom_point(size = 1, shape = 16)+
  #geom_density_2d(color = "green")+
#  labs(subtitle = "PCoA | FKM Clustering | k=3")+
  xlab("Axis 1 [15.47 %]")+ # find out about the percentage by "summary(pcoa.obj)"
  ylab("Axis 2 [9.12 %]")+
  scale_color_viridis(begin = 0, end= 1, option = "inferno", name = "Classification\nStrength")+
  #scale_color_gradientn(colors = c("#111962", "#193F90", "#B65417", "orange", "#EBCF3E"), name = "classification \nstrength")+
  theme(legend.position = c(0.1,0.75),
        axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank(),
        axis.title = element_text(size = 10)
        #plot.margin = margin(2,0,0,0, unit = "cm")
        )
  
pprob
```



# ET-dysbiosis Score by ET
```{r, fig.height = 3, fig.width=7}

ptest <- metadata %>% 
  group_by(fkm_k3) %>% summarize(p.value = kruskal.test(ET_dysbiosis ~ is.diseased )$p.value) %>% 
  mutate(signif = case_when(between(p.value, 0.05, 0.01) ~ "*", 
                            between(p.value, 0.01, 0.001) ~ "**", 
                            p.value < 0.001 ~ "***"))
ptest 

p.et.dys.single <- metadata %>% 
 filter(is.diseased != "unknown disease status") %>% droplevels() %>% 
  ggplot(aes(y=ET_dysbiosis, x = fkm_k3))+
 geom_flat_violin(aes(fill = is.diseased), position = position_nudge(x=.2, y=0), alpha = 0.5)+
  geom_boxplot(width = 0.2, color= "grey20", alpha = 0.5,position= position_dodge(width = 0.2), aes(fill = is.diseased))+
  scale_fill_manual(values = disease_colors)+
  theme(legend.position =  "none")+
 geom_text(data = ptest, aes(x = c(1,2,3), y = c(1.01,1.01,1.01), label = signif))+

  xlab("Enterortype")+
  ylab("")


p2 <- metadata %>% 
  filter(is.diseased != "unknown disease status") %>% 
  ggplot(aes(x=ET_dysbiosis, y = is.diseased, fill = is.diseased))+
  geom_boxplot(alpha = 0.5)+
  theme_publication()+
  scale_fill_manual(values = disease_colors)+
  stat_compare_means(comparisons= list(c("healthy subjects", "patients")), aes(label = after_stat(p.signif)), vjust = 0.4)+
  xlab("Enterotype based Dysbiosis Score")+ylab("")+
  theme(legend.position = c(2.3, 0.14),
        legend.background = element_blank(),
        legend.key.size = unit(0.5, "cm"),
        legend.text = element_text(size = 8),
        legend.title = element_blank())+
  coord_flip()

```


# ETD for each disease
```{r}
disease_dictionary <- list(
  "T2D" = "Type 2 Diabetes",
  "CD" = "Crohn's Disease",
  "ADA" = "Advanced Colorectal Adenoma",
  "CRC" = "Colorectal Cancer",
  "T1D" = "Type 1 Diabetes",
  "RA" = "Rheumatoid Arthritis",
  "PD" = "Parkinson's Disease",
  "NAA" = "Non advanced adenoma",
  "VKH" = "Vogt-Koyanagi-Harada Disease",
  "IGT" = "Impaired glucose tolerance",
  "liver cirrhosis" = "Liver Cirrhosis", 
  "breast cancer" = "Breast Cancer", 
  "end-stage renal disease" = "End-stage Renal disease",
  "atherosclerotic cardiovascular disease" = "Atherosclerotic Cardiovascular disease", 
  "gestational diabetes mellitus" = "Gestational diabetes mellitus",
  "myalgic encephalomyelitis/chronic fatigue syndrome" = "chronic fatigue syndrome", 
  "schizophrenia" = "Schizophrenia", 
  "chronic fatigue syndrom" = "Chronic Fatigue Syndrom",
  "ankylosing spondylitis" = "Ankylosing Spondylitis",
  "UC" = "Ulcerative colitis"
)

metadata <- metadata %>% 
  mutate(disease = recode(disease, !!!disease_dictionary))
EDS_healhty <- metadata %>% 
  filter(is.diseased == "healthy subjects") %>% 
  pull(ET_dysbiosis)
EDS_healthy_median <- EDS_healhty %>% 
  median()


perform_test <- function(disease_values) {
  wilcox.test(disease_values, EDS_healhty)$p.value
}

p_values <- tapply(metadata$ET_dysbiosis, metadata$disease, perform_test) %>% 
  as.data.frame() %>% magrittr::set_colnames("p.value") %>% 
  rownames_to_column("disease") %>% 
  mutate(p.adj = p.adjust(p.value,method =  "BH"))
p_values$stars <- cut(p_values$p.adj, breaks=c(-Inf, 0.001, 0.01, 0.05, Inf), label=c("***", "**", "*", "")) 

n_samples <- metadata %>%
  count(disease) %>% 
  filter(!is.na(disease))

plot_each_disease <- metadata %>% 
  group_by(disease) %>% 
  filter(!is.na(disease)) %>% 
  mutate(EDS_mean_disease = mean(ET_dysbiosis)) %>% 
  arrange(EDS_mean_disease) %>% 
  mutate(disease = factor(disease, levels = unique(.$disease))) %>% 
  ggplot(aes(x=ET_dysbiosis, y = disease))+
  geom_boxplot(alpha = 0.5, aes(fill = is.diseased))+
  geom_vline(xintercept = EDS_healthy_median, color = "#ef7a1a")+
  geom_text(data = p_values, aes(x=0, y = disease, label = stars))+
  theme_publication()+
  scale_fill_manual(values = disease_colors)+
  annotate("text", x =0.35, y = 33,  size = 2, label = "meadian healthy subjects", color = "#ef7a1a")+
  xlab("Enterotype based Dysbiosis Score")+ylab("")+
  theme(legend.position = "none")+
  geom_text(data = n_samples, aes(label = paste("n=", n), y = disease), x = 1.05, size =2, color = "black")+
  xlim(0, 1.1)

       
plot_each_disease
 
```


# ET-Dysbiosis vs. Lloyd-Dysbiosis


```{r fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
 
metadata %>%  
  ggplot(aes(x=activity_index, y = ET_dysbiosis))+
  geom_point(alpha = 0.03, color = "grey50")+
  geom_smooth(color = "black", method = "lm")+
  geom_density_2d()+
  ggpubr::stat_cor(color="black", label.y.npc = 0.01, size = 2)+
    xlab(" ")+
  ylab("")+
  labs(subtitle = "All Samples")+
  theme_publication()

metadata %>% 
  filter(fkm_k3 == "Prevotella") %>% 
  ggplot(aes(x=activity_index, y = ET_dysbiosis))+
  geom_point(alpha = 0.03, aes(color = fkm_k3))+
  geom_smooth(color = "black", method = "lm")+
  ggpubr::stat_cor(color="black", label.y.npc = 0.01, size = 2)+  
    theme_publication()+
  theme(legend.position = "none")+
  scale_color_manual(values = enterotype_colors)+
  xlab("")+ylab("")+labs(subtitle ="Prevotella")

metadata %>% 
  filter(fkm_k3 == "Bact_Phoc") %>% 
  ggplot(aes(x=activity_index, y = ET_dysbiosis))+
  geom_point(alpha = 0.03, aes(color = fkm_k3))+
  geom_smooth(color = "black", method = "lm")+
  ggpubr::stat_cor(color="black", label.y.npc = 0.01, size = 2)+  
    theme_publication()+
  theme(legend.position = "none")+
  scale_color_manual(values = enterotype_colors)+
  xlab("")+ylab("")+labs(subtitle = "Bact_Phoc")

metadata %>% 
  filter(fkm_k3 == "Firmicutes") %>% 
  ggplot(aes(x=activity_index, y = ET_dysbiosis))+
  geom_point(alpha = 0.03, aes(color = fkm_k3))+
  geom_smooth(color = "black", method = "lm")+
  ggpubr::stat_cor(color="black", label.y.npc = 0.01, size = 2)+  
    theme_publication()+
  theme(legend.position = "none")+
  scale_color_manual(values = enterotype_colors)+
  xlab("")+ylab("")+labs(subtitle= "Firmicutes")


```

## correlation with meldScore 
```{r}
qin.data <- metadata %>% filter(studies == "Qin_2014_liver_cirrhosis")
qin.extra <- readRDS(here("data/Qin_2014_meld.rds"))


Qin <- qin.data %>% left_join(qin.extra, by = c("internal_id"="ena_ers_sample_id"))

qin.correlation.plot <- Qin %>% 
  ggplot(aes(x=ET_dysbiosis, y= as.numeric(MELD)))+
  geom_point()+
  stat_cor(method="spearman")+
  geom_smooth(method = "lm", color = "grey20")+
  xlab("Enterotype Dysbiosis Score (EDS)")+ylab("Liver Disease Severity [meld score]")
qin.correlation.plot
```

# ET-Dysbiosis vs. Shannon 


```{r  fig.width = 10, fig.height  = 3}

metadata %>%  
  ggplot(aes(x=Shannon, y = ET_dysbiosis))+
  geom_point(alpha = 0.3, color = "grey30")+
#  facet_wrap(~Cluster_k3,ncol=2 )+
  geom_smooth(color = "black", method ="lm")+
  geom_density_2d()+
  ggpubr::stat_cor(color="black")+
    xlab("")+
  ylab("Enterotype based Dysbiosis Score")+
  labs(subtitle = "All Samples")

```


# EDS in RIFSYS cohort 

the sample-wise clinical data for the RIFSYS cohort is accessible for the authors through the MICROB-PREDICT consortium but cannot be made publicly available due to general data protection laws (GDPR). Please refer to the original publication for accessing the clinical data. 

```{r eval=FALSE, include=FALSE}
et <- read_tsv(here("manuscript/results/EnterotypeAssignment_fuzzyClustering_k3_Patel_2020_RIFSYS.tsv"))
data <- et %>% left_join(ids, by =c("sample_id" = "Sample")) %>% 
  left_join(clinData, by ="ID")
```


Correlation with change in Neutrophil Oxidative Burst
```{r eval=FALSE, include=FALSE}
rifsys.plot <- data %>%  
  filter(!is.na(GROUP) & timepoint == "d01") %>% 
  mutate(Log2ChangeNtrphlBrst30 = log2(Out30NtrphlBrst/OutNtrphlBrst)) %>% 
  ggplot(aes(x=Enterotype_Dysbiosis_Score, y = Log2ChangeNtrphlBrst30))+
  geom_point()+
  geom_smooth(method = "lm")+
  stat_cor(method = "kendall")

rifsys.single.plot <- data %>%  
  filter(!is.na(GROUP) & timepoint == "d01") %>% 
  mutate(Log2ChangeNtrphlBrst30 = log2(Out30NtrphlBrst/OutNtrphlBrst)) %>% 
  ggplot(aes(x=Enterotype_Dysbiosis_Score, y = Log2ChangeNtrphlBrst30))+
  geom_point()+
  geom_smooth(method = "lm")+
  stat_cor(method = "kendall")+
  facet_wrap(~GROUP)
```


# EDS in GALAXY cohort

the clinical data and the outcome for the GALAXY cohort are accessible to the authors through the GALAXY/MicrobLiver consortium but cannot be made publicly available due to general data protection laws (GDPR).
The script which was used to calculate and visualize the results is shown below

## Difference between people progressing and stable?

```{r, fig.width = 4, fig.height=5}
# outcome file is not available due to GDPR

meta.galaxy <- metadata.galaxy %>% filter(studies == "galaxy") %>% distinct() %>% 
  mutate(id = as.numeric(str_remove_all(.$internal_id, "ALD"))) %>% 
  filter(!is.na(id)) # remove al HP
table(meta.galaxy$id %in% outcome$id)
meta.galaxy <- left_join(meta.galaxy, outcome)

meta.galaxy <- meta.galaxy %>% 
  filter(!is.na(fu_te_cat)) %>% 
  mutate(progression = case_when(fu_te_cat %in% c("Improve", "Stable") ~ "no", 
                                 fu_te_cat == "Progresses" ~ "yes", 
                                 TRUE ~ as.character(NA)))
  
  
galaxy.plot <- meta.galaxy %>% 
  ggplot(aes(x=progression, y = ET_dysbiosis))+
  geom_flat_violin(aes(fill = progression), position = position_nudge(x=.2, y=0), alpha = 0.5)+
  geom_boxplot(width = 0.2, color= "grey20", alpha = 0.5,position= position_dodge(width = 0.2), aes(fill = progression))+
  scale_fill_manual(values = c("#ffce1f", "#aa3e98"))+
  stat_compare_means(comparisons = list(c("yes", "no")), 
                     method = "wilcox.test",
                     label = "p.signif")+
  theme(legend.position = "none")+
  xlab("Disease progression")+ylab("Enterotype Dysbiosis Score")

```


## Kaplan Meier for GALAXY death

```{r fig.height = 6, fig.width = 6}
library(survival) 
library(survminer)

galaxy.km <- meta.galaxy %>% 
  filter(censoring_reason != "Lost to follow-up") %>% 
  select(id,subject_id, internal_id, ET_dysbiosis, incl, censoring_date, death) %>% distinct() %>% 
  mutate(survival_time = censoring_date - incl, 
         event = case_when(death == "Yes" ~ 1, 
                           death == "No" ~ 0, 
                           TRUE ~NA )) %>% 
     mutate(is.dysbiosis = case_when(ET_dysbiosis >= 0.77 ~ "high", 
                                  ET_dysbiosis < 0.77 ~ "low")) 

# with same sized groups (using median of ET_Dysbiosis in the ALD cohort as cutoff)

treshhold.equal <- median(meta.galaxy$ET_dysbiosis)
galaxy.km <- galaxy.km %>% 
 mutate(dysbiosis.group = case_when(ET_dysbiosis >= treshhold.equal ~ "high", 
                                  ET_dysbiosis < treshhold.equal ~ "low")) 

  
surv_object <- Surv(time = galaxy.km$survival_time, event = galaxy.km$event)

fit1 <- survfit(surv_object ~ dysbiosis.group, data = galaxy.km)
kaplan.dysbiosis <- ggsurvplot(fit1, data = galaxy.km, pval = TRUE ,conf.int=TRUE, combine = TRUE, risk.table=TRUE, palette = c("#723c70","#E9C46A"))
kaplan.dysbiosis$plot <- kaplan.dysbiosis$plot + theme(legend.position = "none")
kaplan.dysbiosis

```

## Kaplan Meier with EDS and TE 
```{r, fig.width=8, fig.height=8}

# with cutoff to have equal sized groups (meadian of TE values) 
treshhold.equal.te <- median(galaxy.km.clinical$te, na.rm = TRUE) # median is 7.35
galaxy.km.clinical <- galaxy.km.clinical %>% 
 mutate(te.group = case_when(te >= treshhold.equal.te ~ "high", 
                                  te < treshhold.equal.te ~ "low"))

  
surv_object <- Surv(time = galaxy.km.clinical$survival_time, event = galaxy.km.clinical$event)

fit1 <- survfit(surv_object ~ te.group + dysbiosis.group, data = galaxy.km.clinical)
kaplan.dysbiosis <- ggsurvplot(fit1, data = galaxy.km.clinical, pval = TRUE ,conf.int=FALSE, risk.table=TRUE, palette = c("#B5838D", "#723c70", "#E9C46A", "#ffa62b"))
kaplan.dysbiosis$plot <- kaplan.dysbiosis$plot + theme(legend.position = "none")
kaplan.dysbiosis

```

# EDS in MICROB-PREDICT cohort 

The time to death cannot be publicly available due to GDPR regulations.

## Kaplan Meier with EDS
```{r}
md.public <- read_tsv(here("data/MICROB-PREDICT_sample_data.tsv")) 
treshhold.equal <- median(md.public$ET_dysbiosis)
md.public <- md.public %>% 
  mutate(is.dysbiosis.equal = case_when(ET_dysbiosis >= treshhold.equal ~ "high", 
                                        ET_dysbiosis <= treshhold.equal ~ "low"))

# surv_object <- Surv(time = md.public$time_to_death, event = md.public$mort_90d)

fit1 <- survfit(surv_object ~ is.dysbiosis.equal, data = md.public)
kaplan.dysbiosis <- plot <- ggsurvplot(fit1, data = md.public, pval = TRUE ,conf.int=TRUE, risk.table=TRUE, palette = c( "#723c70", "#E9C46A"))
kaplan.dysbiosis$plot <- kaplan.dysbiosis$plot + theme(legend.position = "none")
kaplan.dysbiosis
```

## Kaplan Meier with EDS and MELD
```{r}
treshold.meld <-median(md.public$meld)
treshold.eds <-median(md.public$ET_dysbiosis)
md.public <- md.public %>% 
  mutate(meld.group = case_when(meld >= treshold.meld ~ "high", 
                                  meld < treshold.meld ~ "low"), 
         dysbiosis.group = case_when(ET_dysbiosis >= treshold.eds ~ "high", 
                                     ET_dysbiosis < treshold.eds ~ "low")) 


surv_object <- Surv(time = md.public$time_to_death, event = md.public$mort_90d)

fit1 <- survfit(surv_object ~ meld.group + dysbiosis.group, data = md.public)

kaplan.dysbiosis <- ggsurvplot(fit1, data = md.public, pval = TRUE ,conf.int=FALSE, risk.table=TRUE, palette = c("#B5838D", "#723c70", "#E9C46A", "#ffa62b"), legend = "right")
kaplan.dysbiosis$plot <- kaplan.dysbiosis$plot + theme(legend.position = "none")
kaplan.dysbiosis
```

# Linear Model Taxa ~ EDS

## all samples
```{r}
data_adults = counts
# packages
library(compositions)

metadata = metadata %>% 
  column_to_rownames("internal_id")
# align samples
common <- intersect(colnames(data_adults), rownames(metadata))
X <- data_adults[, common, drop=FALSE]
meta <- metadata[common, , drop=FALSE]
meta = metadata %>% 
  select(studies, fkm_k3, ET_dysbiosis, disease)
# CLR transform (add small pseudocount)
X_clr <- t(clr(t(X + 1e-6)))  # keep genera x samples; clr expects samples x features so we transpose twice
X_clr = as.matrix(X_clr)
# choose covariates you have (edit this list!)
covars <- c("studies", "disease")

# build formula string
rhs <- paste(c("ET_dysbiosis", covars), collapse = " + ")

# fit genus-wise linear models: clr(genus) ~ ET + covariates
res <- lapply(rownames(X_clr), function(g) {
  df <- data.frame(y = as.numeric(X_clr[g, ]), meta)
  fit <- lm(as.formula(paste("y ~", rhs)), data=df)

  s <- summary(fit)$coefficients
  if (!"ET_dysbiosis" %in% rownames(s)) return(NULL)

  data.frame(
    genus = g,
    beta = s["ET_dysbiosis","Estimate"],
    se   = s["ET_dysbiosis","Std. Error"],
    t    = s["ET_dysbiosis","t value"],
    p    = s["ET_dysbiosis","Pr(>|t|)"],
    row.names = NULL
  )
})

res <- do.call(rbind, res)
res$padj <- p.adjust(res$p, method="BH")
res <- res[order(res$padj, res$p), ]
head(res, 20)
res %>% filter(padj>0.05)

saveRDS(res, here("fulldata_v2/analysis/adults_filtered_gtdb/dysbiosisScore/LM_EDS.rds"))
```

## only Prevotella samples

```{r}
# filter metadata for prevotella samples
meta_prev = metadata %>% filter(fkm_k3 == "Prevotella")  

# align samples
common <- intersect(colnames(data_adults), rownames(meta_prev))
X_prev <- data_adults[, common, drop=FALSE]
meta_prev <- meta_prev[common, , drop=FALSE]
meta_prev = meta_prev %>% 
  select(studies, fkm_k3, ET_dysbiosis, disease)
# CLR transform (add small pseudocount)
X_prev_clr <- t(clr(t(X_prev + 1e-6)))  # keep genera x samples; clr expects samples x features so we transpose twice
X_prev_clr = as.matrix(X_prev_clr)
# choose covariates you have (edit this list!)
covars <- c("studies", "disease")

# build formula string
rhs <- paste(c("ET_dysbiosis", covars), collapse = " + ")

# fit genus-wise linear models: clr(genus) ~ ET + covariates
res_prev <- lapply(rownames(X_prev_clr), function(g) {
  df <- data.frame(y = as.numeric(X_prev_clr[g, ]), meta_prev)
  fit <- lm(as.formula(paste("y ~", rhs)), data=df)

  s <- summary(fit)$coefficients
  if (!"ET_dysbiosis" %in% rownames(s)) return(NULL)

  data.frame(
    genus = g,
    beta = s["ET_dysbiosis","Estimate"],
    se   = s["ET_dysbiosis","Std. Error"],
    t    = s["ET_dysbiosis","t value"],
    p    = s["ET_dysbiosis","Pr(>|t|)"],
    row.names = NULL
  )
})

res_prev <- do.call(rbind, res_prev)
res_prev$padj <- p.adjust(res_prev$p, method="BH")
res_prev <- res_prev[order(res_prev$padj, res_prev$p), ]
head(res_prev, 20)
res_prev %>% filter(padj>0.05)
saveRDS(res, here("fulldata_v2/analysis/adults_filtered_gtdb/dysbiosisScore/LM_EDS_Prevotella.rds"))
```

## only Firmicutes samples

```{r}
# filter metadata for prevotella samples
meta_firm = metadata %>% filter(fkm_k3 == "Firmicutes")  

# align samples
common <- intersect(colnames(data_adults), rownames(meta_firm))
X_firm <- data_adults[, common, drop=FALSE]
meta_firm <- meta_firm[common, , drop=FALSE]
meta_firm = meta_firm %>% 
  select(studies, fkm_k3, ET_dysbiosis, disease)
# CLR transform (add small pseudocount)
X_firm_clr <- t(clr(t(X_firm + 1e-6)))  # keep genera x samples; clr expects samples x features so we transpose twice
X_firm_clr = as.matrix(X_firm_clr)
# choose covariates you have (edit this list!)
covars <- c("studies", "disease")

# build formula string
rhs <- paste(c("ET_dysbiosis", covars), collapse = " + ")

# fit genus-wise linear models: clr(genus) ~ ET + covariates
res_firm <- lapply(rownames(X_firm_clr), function(g) {
  df <- data.frame(y = as.numeric(X_firm_clr[g, ]), meta_firm)
  fit <- lm(as.formula(paste("y ~", rhs)), data=df)

  s <- summary(fit)$coefficients
  if (!"ET_dysbiosis" %in% rownames(s)) return(NULL)

  data.frame(
    genus = g,
    beta = s["ET_dysbiosis","Estimate"],
    se   = s["ET_dysbiosis","Std. Error"],
    t    = s["ET_dysbiosis","t value"],
    p    = s["ET_dysbiosis","Pr(>|t|)"],
    row.names = NULL
  )
})

res_firm <- do.call(rbind, res_firm)
res_firm$padj <- p.adjust(res_firm$p, method="BH")
res_firm <- res_firm[order(res_firm$padj, res_firm$p), ]
head(res_firm, 20)
res_firm %>% filter(padj>0.05)

saveRDS(res, here("fulldata_v2/analysis/adults_filtered_gtdb/dysbiosisScore/LM_EDS_Firmicutes.rds"))
```



## only Bacteroides / Phocaeicola samples

```{r}
# filter metadata for prevotella samples
meta_bact = metadata %>% filter(fkm_k3 == "Bact_Phoc")  

# align samples
common <- intersect(colnames(data_adults), rownames(meta_bact))
X_bact <- data_adults[, common, drop=FALSE]
meta_bact <- meta_bact[common, , drop=FALSE]
meta_bact = meta_bact %>% 
  select(studies, fkm_k3, ET_dysbiosis, disease)
# CLR transform (add small pseudocount)
X_bact_clr <- t(clr(t(X_bact + 1e-6)))  # keep genera x samples; clr expects samples x features so we transpose twice
X_bact_clr = as.matrix(X_bact_clr)
# choose covariates you have (edit this list!)
covars <- c("studies", "disease" )

# build formula string
rhs <- paste(c("ET_dysbiosis", covars), collapse = " + ")

# fit genus-wise linear models: clr(genus) ~ ET + covariates
res_bact <- lapply(rownames(X_bact_clr), function(g) {
  df <- data.frame(y = as.numeric(X_bact_clr[g, ]), meta_bact)
  fit <- lm(as.formula(paste("y ~", rhs)), data=df)

  s <- summary(fit)$coefficients
  if (!"ET_dysbiosis" %in% rownames(s)) return(NULL)

  data.frame(
    genus = g,
    beta = s["ET_dysbiosis","Estimate"],
    se   = s["ET_dysbiosis","Std. Error"],
    t    = s["ET_dysbiosis","t value"],
    p    = s["ET_dysbiosis","Pr(>|t|)"],
    row.names = NULL
  )
})

res_bact <- do.call(rbind, res_bact)
res_bact$padj <- p.adjust(res_bact$p, method="BH")
res_bact <- res_bact[order(res_bact$padj, res_bact$p), ]
head(res_bact, 20)
res_bact %>% filter(padj>0.05)

saveRDS(res, here("fulldata_v2/analysis/adults_filtered_gtdb/dysbiosisScore/LM_EDS_Bacteroides.rds"))
```




